---
title: "MDSS Introduction to R Workshop Part 2"
subtitle: "Basics of R"
author: "Joshua Mills"
length: "50 minutes"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
    theme: lumen

---


# Data Frames - Loading and manipulating data

## Installing and loading packages

For the next section we are going to be using the *readr*, *tidyr* and *dplyr* packages. They provide many useful functions and data types that allow us to manipulate and read data. To install a package use the following commands.

If you have not installed these packages from last session then these commands will.

```{r eval=FALSE}
install.packages("dplyr")
install.packages("readr")
install.packages("ggplot2")
install.packages("tidyr")
```

We will first load the packages we have just installed.

```{r}
library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
```

Data frames are a core part of the R language, almost all applications of R use data frames to store and manipulate data.

## Tidy Data

When dealing with data frames in R we typically try and keep the data in a tidy format. There are essentially 3 rules for tidy data: 

1. Each variable forms a column.

2. Each observation forms a row.

3. Each type of observational unit forms a table.

Data in tidy form is a lot easier to analyse.

The packages we are going to be using in the rest of the workshop are from a collection called the tidyverse. 
(https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html)


## Data Frames

In R a lot of the data we store is in Data frames. These are similar to using a single spreadsheet in excel, but R has many fantastic functionalities that we can use to sort, select, summarize and produce graphics, statistical analyses  and machine learning from data frames. 
We will start by reading in a csv file that contains records of Melbourne's Housing sales. 

Before we can start manipulating/analyzing data and extracting any insights, we have to read it into R. We may use the function `read_csv` from *readr* to read data directly from the web. It takes a CSV file and converts into a data frame that we can manipulate.

```{r}
HousingData <- readr::read_csv("https://raw.githubusercontent.com/joshuajohnmills/Rworkshop/master/MelbourneHousingData.csv")
```

**Viewing the dataframe**
In order to view a data frame in Rstudio we can use the View() function. Do not use this function if you are going to work in a different IDE. If using a downloaded version of R, you can use the `fix()` function to directly edit a dataframe. 

```{r eval=FALSE}
View(HousingData)
```

## The Magrittr pipe

The Magrittr pipe is an operator in R that allows us to do multiple functions in one. It takes the previous variable and puts it into the first argument in the next function. For example

```{r}
16 %>% log(2) %>% sum(3,2) %>% sqrt()
#takes the log of 16 and adds  it to 3 and 2, then takes the square root it is the same as the next line 
```

is exactly the same as 

```{r}
sqrt(sum(log(16,2),3,2))
```

but a lot easier to organize.This might seem trivial for integers, but when working with data frames it is very useful. 


## Simple Data Manipulation


### Filter and arrange 

`filter` is a verb in *dplyr* which lets us filter for rows that satisfy a certain constraint. it allows us to see subsets of the data frame that we are interested in.

`arrange` sorts the dataset by the column that is specified in the arguments. 

`desc` means makes the sorting in descending order.

Note that here we first select then use the *magrittr pipe* to pass it on to `arrange`. It will automatically execute. 

```{r}
HousingData %>% filter(Suburb == "Caulfield") %>% arrange(desc(Price))
```


**Questions**

1. Find the highest price in your suburb.

2. List the house with the most bedrooms.

3. What is the most recent sale date in the dataset?

4. Which one is the oldest house in the dataset?

The tidyverse is a collection of some of the most popular and useful libraries in R. It contains too many libaries and functionalities to be covered in only a few workshops, so we will be looking at a few of the more general purpose ones. For this session, we will use *readr* for reading in data and *dplyr* for wrangling and reshaping data and in session 3, *ggplot* to create graphs and plots. 

# Getting Started with *dplyr*

*dplyr* is useful for manipulating data frames and wrangling data to get insights and to prepare it for visualisation and machine learning. It contains multiple useful functions. (Some of these functions are the same as SQL query keywords)

## Select 

Select takes specified columns from a data frame. Here we are only looking for columns that are specified. Say for example we want to find out what suburbs have the cheapest (or most expensive) houses. We can then select the suburbs and the price of each house that has sold. 

```{r}
HousingPrices <- HousingData %>% 
  select(Suburb,Price,Bedroom2) 
```

## Filter

Just like `select` chooses columns from dataframe, `filter` can be used to choose rows. If we are only interested in houses with two or less bedrooms we can use `filter` to specify the criteria for choosing only the rows where bedrooms are less than or equal to two (bedroom <= 2).

```{r}
HousingBedrooms <- HousingPrices %>% 
  filter(Bedroom2 <= 2)
#HousingBedrooms %>% head(10) %>%  knitr::kable()
```

This will return houses that have less than three bedrooms and store that in `HousingBedrooms` dataframe. To view the resultant dataframe, `View()` function can be used.

**Question**

1. Find all houses that were recorded before 1900 (Date is before 1900).

## Group by and summarise 

Say we want to find out what suburb is the most expensive. We can go over the data for each suburb (group) and calculate average house price per suburb. 

To achieve this here, we first specify what variable we want to group by, and then we can summarise each group to get the average Price.

```{r}
HousingData %>% group_by(Suburb) %>% summarise(mean(Price))
```

The order of dataframe is preserved by this command, but this may not be very useful. There are many many suburbs so it might be silly to look through each row to find the maximums. The `arrange` function can solve this:

```{r}
HousingData %>% group_by(Suburb) %>% summarise(Price=mean(Price)) %>% arrange(desc(Price))
```

Use `Count=n()` to find the amount of houses sold in each suburb:

```{r}
HousingData %>% group_by(Suburb) %>% summarise(Price = mean(Price),Count= n()) %>% arrange(desc(Price))
```

## Mutate 

`mutate` function creates a new column.
Say we want to find the average price per room for each council area counting, say we count a bedroom as 1 room and a bathroom as half a room.
`mutate` will add a new column called `PricePerRoom` which is equal to the price divided by the number of bedrooms plus half the number of bathrooms. 

```{r}
HousingData %>%
  mutate(PricePerRoom = Price/(Bedroom2 + 0.5*Bathroom))
```

Putting this together to find the council areas that have the most expensive price per room.

```{r}
HousingData %>%
  mutate(PricePerRoom = Price/(Bedroom2 + 0.5*Bathroom)) %>% 
  group_by(CouncilArea) %>% 
  summarise(PricePerRoom=median(PricePerRoom),count=n()) %>% 
  arrange(desc(PricePerRoom))
```

**Question:**

1. Find suburbs that have the oldest house(s).

##Dealing with Na's (Missing Values)

`drop_na()` will drop all columns with na in the value.

```{r}
AgesHousing <- HousingData %>% drop_na(YearBuilt)
```


**Questions:**

1. Create a data frame that only has houses with land area above 50sqm.

2. Find the cost per sqm in each suburb.









